Convergence rate of MAP estimates \\ for the exponential family

### 2020-12-14 main.tex
Minor changes

The exponential family member with sufficient statistic $T$ and natural parameter $\natp$ is the model
\begin{align}
    p(X|\natp) &= \exp( \natp^\top T(X) - \logpart(\natp)) \; ,
\end{align}
with log-partition function 
\begin{align}
    \logpart(\natp) = \log \int e^{\natp^\top T(x)} dx \; .
\end{align}
Recall that $\logpart$ verifies the two following identities 
\begin{align}
    \nabla\logpart(\natp) &=  \expect[p(X|\natp)]{T(X)} =: \meanp \\
    \nabla^2 \logpart(\natp) &= \Cov_\natp[T(X)] > 0
\end{align}
where $\meanp$ is called the mean parameter.
The second identity entails that $\logpart$ is strictly-convex.
The conjugate prior for this distribution is
\begin{align}
    \pi(\natp) &\propto \exp( - n_0 \bregman(\natp || \natp_0) )
\end{align}
where $n_0$ is a number of fictional points observed from a distribution with parameter $\natp_0$.
$\bregman(\natp || \natp_0)$ is the Bregman divergence induced by $\logpart$ between $\natp$ and $\natp_0$
\begin{align}
    \bregman (\natp || \natp_0)
    & = \logpart(\natp) - \logpart(\natp_0) 
    - \langle \meanp_0 , \natp - \natp_0 \rangle
\end{align}
with $\meanp_0 = \nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)}$ the mean parameter associated to the natural parameter $\natp_0$.
The negative log-likelihood of the prior is then
\begin{align*}
    -\log \pi(\natp) = n_0 (\logpart(\natp)  - \natp^\top \meanp_0 ) + \cst
\end{align*}
Thus the joint NLL of $(x_1,\dots,x_n,\natp)$ is
\begin{align}
    -\log p(X|\natp)\pi(\natp) 
    = (n_0+n) \logpart (\natp) 
    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T(x_i) \right) \; .
\end{align}
Minimizing this expression over $\natp$ yields the Maximum A Posteriori estimate
\begin{align}
    \hat \natp = \argmin_\natp -\log p(X|\natp) + n_0 \bregman(\natp || \natp_0)
\end{align}
such that the MAP is
\begin{align}
    \nabla \logpart(\hat \natp) = \hat \meanp
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T(X_i) }{n_0+n} \; .
\end{align}
The MAP estimate is a random quantity.

\section{Straightforward Convergence Rate}
In the realizable case, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{align}
    &\expect[x\sim p(.|\natp^*)]{-\log p(x|\natp) + \log p(x|\natp^*)}\\
    =& \KL( p(.|\natp^*) || p(.|\natp)) \\
    =& \bregman (\natp || \natp^*) \\
    =& \bregmanconj ( \meanp^* || \meanp) \\
    =& \conj(\mu^*) + \logpart(\natp) - \langle \natp , \mu^* \rangle
    %& = \logpart(\natp) - \logpart(\natp^*) - \meanp^{*\top}(\natp - \natp^*) \; .
\end{align}
where $\conj$ is the entropy, the convex conjugate of the log-partition.
The relationship between Bregman divergences and Fenchel conjugacy is well explained in \citet{wainwright2008graphical}, and \citet{agarwal2010geometric}. 
The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds -- in expectation or high-probability ?
\begin{align}
	\label{eq:bregmanMLE}
	\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(x_i) \right ) \\
	\label{eq:bregmanMAP}
	\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right ) 
\end{align}

@@

This is very interesting for us, but the variance hyper-parameter is oddly defined. Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate. Then the variance bound $\sigma^2$ is an upper bound on the covariance between the gradient update  $-g_t$ and the descent direction $\natp_{t+1} - \natp_t = \nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t g_t ) - \natp_t $ that should hold for all time steps  
\begin{align}
	&\Cov(- g_t,\natp_{t+1} - \natp_t | X_{1 \dots t}) \leq \lr_t \sigma^2 \\
	&= \expect{\langle - g_t, \natp_{t+1} \rangle}  - \langle \E[-g_t], \E[\natp_{t+1}] \rangle   \\
	 &= \expect{\langle \nabla f(\natp_t) - g_t , \natp_{t+1}  \rangle}
\end{align}
where $\gamma_t$ is the step-size and expectations are conditional on the past. Remark that when we plug in $\conj = \|.\|^2$, we recover the gradient variance typical of SGD. In our case, 
\begin{align}
	\nabla f(\natp_t) - g_t =  T(X_{t+1})	- \E[T(X)]
\end{align}
so that $\sigma$ is really a bound on the covariance of the sufficient statistics with another variable
\begin{align}
	&\Cov( T(X_{t+1}), \natp_{t+1}| X_{1\dots t})
	\leq \lr_t\sigma^2 \\
	=& \Cov(T(X_{t+1}), \nabla\conj \left(\lr_t T(X_{t+1}) + (1-\lr_t ) \mu_t \right) | X_{1\dots t}) \;.
\end{align}
In other words, we need for all $\lr\in(0,1)$, and all $\mu$ in the interior of the marginal polytope,
\begin{align}
\label{eq:hypvariance}
	\Cov_X \bigg(T(X), \nabla \conj (\lr T(X) + (1-\lr)\mu)\bigg	) 
	\leq \lr \sigma^2 \; .
\end{align}
When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, assuming all $T(x)$ belong to the marginal polytope, we get
\begin{multline}
	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ||\hat \mu) }
	=\conj(\mu_*) \overbrace{- \E[\conj(\hat \mu)]}^{\leq - \conj(\E[\hat \mu])} \\
	+ \inv{n}\sum_i \E[\underbrace{\E[\langle  T(X_i) - \mu_*; \nabla\conj (\hat \mu) \rangle | X_j, j\neq i]}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance} } ]  \leq \frac{\sigma^2}{n}
\end{multline}
where we used the decomposition $\hat \mu = \inv{n}T(X_i) + (1-\inv{n}) \inv{n-1} \sum_{j\neq i}T(X_j)$ to apply~\eqref{eq:hypvariance}. 
In words, this variance assumption on $T(X)$ and $\conj$ immediately gives us a bound on the suboptimality. We can also apply this result to the MAP estimate $\hat \mu = \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n}$ but I did not manage to reach a satisfying conclusion about the bias. 
Note $\tilde \mu = \E[\hat \mu] = \frac{n_0 \mu_0 + n \mu_*}{n_0+n}$ and $\gamma_0= \frac{n_0}{n_0+n}$ and the step-size $\gamma = \inv{n_0 + n}$ .
\begin{align}
	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ||\hat \mu) } 
	&\leq \conj(\mu_*) - \conj(\tilde \mu) + \gamma_0 \langle \mu_0 - \mu_* ; \E[\hat \natp] \rangle + (1 - \gamma_0) \gamma \sigma^2  \\
	& = \bregmanconj (\mu_* ; \tilde \mu) - \langle \tilde \mu - \mu_* ;    \tilde \natp - \E[\hat \natp] \rangle + \frac{n}{(n_0+n)^2}\sigma^2
\end{align}
so we recover the $O(n^{-1})$ for the variance and we are still looking for $O(n^{-2})$ rate for the bias. The Bregman term in the bias should be in this spirit, given a quadratic approximation
\begin{align}
	 \bregmanconj (\mu_* ; \tilde \mu)  
	 = \bregmanconj (\mu_* ; \mu_* + \gamma_0 (\mu_0 - \mu_*) ) \approx \half \gamma_0^2 \norm{\mu_0 - \mu_*}^2_{\mSigma_*^{-1}}
\end{align}
where $\mSigma_* = \Cov_{\mu_*}(T(X))$. \RLP{beware of the order, this might be inaccurate}
However the scalar product is harder to bound. Hopefully it could be positive. If not we know that $\tilde \mu - \mu_* = \frac{n_0}{n_0+n} (\mu_0 - \mu_*) = O(n^{-1})$ , and we can hope that the same holds for $\tilde \natp - \E[\hat \natp] = \nabla \conj(\E[\hat \mu]) - \E[\nabla\conj (\hat \mu)]$, which kinda measures the non-linearity, or the curvature of $\conj$. Using a simple cauchy-schwartz, and the hessian of the conjugate, we might be able to get something.

>>>>

The exponential family member with sufficient statistic $T$ and natural parameter $\natp$ is the model
\begin{align}
    p(X|\natp) &= \exp( \natp^\top T(X) - \logpart(\natp)) \; ,
\end{align}
with log-partition function 
\begin{align}
    \logpart(\natp) = \log \int e^{\natp^\top T(x)} dx \; .
\end{align}
Recall that $\logpart$ verifies the two following identities 
\begin{align}
    \nabla\logpart(\natp) &=  \expect[p(X|\natp)]{T(X)} =: \meanp \\
    \nabla^2 \logpart(\natp) &= \Cov_\natp[T(X)] > 0
\end{align}
where $\meanp$ is called the mean parameter.
The second identity entails that $\logpart$ is strictly-convex.
The conjugate prior for this distribution is
\begin{align}
    \pi(\natp) &\propto \exp( - n_0 \bregman(\natp || \natp_0) )
\end{align}
where $n_0$ is a number of fictional points observed from a distribution with parameter $\natp_0$.
$\bregman(\natp || \natp_0)$ is the Bregman divergence induced by $\logpart$ between $\natp$ and $\natp_0$
\begin{align}
    \bregman (\natp || \natp_0)
    & = \logpart(\natp) - \logpart(\natp_0) 
    - \langle \meanp_0 , \natp - \natp_0 \rangle
\end{align}
with $\meanp_0 = \nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)}$ the mean parameter associated to the natural parameter $\natp_0$.
The negative log-likelihood of the prior is then
\begin{align*}
    -\log \pi(\natp) = n_0 (\logpart(\natp)  - \natp^\top \meanp_0 ) + \cst
\end{align*}
Thus the joint NLL of $(x_1,\dots,x_n,\natp)$ is
\begin{align}
    -\log p(X|\natp)\pi(\natp) 
    = (n_0+n) \logpart (\natp) 
    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T(x_i) \right) + \cst \; .
\end{align}
Minimizing this expression over $\natp$ yields the Maximum A Posteriori estimate
\begin{align}
    \hat \natp = \argmin_\natp -\log p(X|\natp) + n_0 \bregman(\natp || \natp_0)
\end{align}
such that the MAP is
\begin{align}
    \nabla \logpart(\hat \natp) = \hat \meanp
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T(X_i) }{n_0+n} \; .
\end{align}
The MAP estimate is a random quantity.

\section{Straightforward Convergence Rate}
In the realizable case, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{align}
    &\expect[X\sim p(.|\natp^*)]{-\log p(X|\natp) + \log p(X|\natp^*)}\\
    =& \KL( p(.|\natp^*) || p(.|\natp)) \\
    =& \bregman (\natp || \natp^*) \\
    =& \bregmanconj ( \meanp^* || \meanp) \\
    =& \conj(\mu^*) + \logpart(\natp) - \langle \natp , \mu^* \rangle
    %& = \logpart(\natp) - \logpart(\natp^*) - \meanp^{*\top}(\natp - \natp^*) \; .
\end{align}
where $\conj$ is the entropy, the convex conjugate of the log-partition.
The relationship between Bregman divergences and Fenchel conjugacy is well explained in \citet{wainwright2008graphical}, and \citet{agarwal2010geometric}. 
The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds -- in expectation or high-probability -- on the following quantities ?
\begin{align}
	\label{eq:bregmanMLE}
	\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(x_i) \right ) \; , \\
	\label{eq:bregmanMAP}
	\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right ) \; .
\end{align}

@@

This is very interesting for us, but the variance hyper-parameter is oddly defined. Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate. Then the variance bound $\sigma^2$ is an upper bound on the covariance between the gradient update  $-g_t$ and the descent direction $\natp_{t+1} - \natp_t = \nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t g_t ) - \natp_t $ that should hold for all time steps  
\begin{align}
	&\Cov(- g_t,\natp_{t+1} - \natp_t | X_{1 \dots t}) \leq \lr_t \sigma^2 \\
	&= \expect{\langle - g_t, \natp_{t+1} \rangle}  - \langle \E[-g_t], \E[\natp_{t+1}] \rangle   \\
	 &= \expect{\langle \nabla f(\natp_t) - g_t , \natp_{t+1}  \rangle}
\end{align}
where $\gamma_t$ is the step-size and expectations are conditional on the past. Remark that when we plug in $\conj = \|.\|^2$, we recover the gradient variance typical of SGD. In our case, 
\begin{align}
	\nabla f(\natp_t) - g_t =  T(X_{t+1})	- \E[T(X)]
\end{align}
so that $\sigma$ is really a bound on the covariance of the sufficient statistics with another variable
\begin{align}
	&\Cov( T(X_{t+1}), \natp_{t+1}| X_{1\dots t})
	\leq \lr_t\sigma^2 \\
	=& \Cov(T(X_{t+1}), \nabla\conj \left(\lr_t T(X_{t+1}) + (1-\lr_t ) \mu_t \right) | X_{1\dots t}) \;.
\end{align}
In other words, we need for all $\lr\in(0,\inv{n})$, and all $\mu$ in the interior of the marginal polytope,
\begin{align}
\label{eq:hypvariance}
	\Cov_X \bigg(T(X), \nabla \conj (\lr T(X) + (1-\lr)\mu)\bigg	) 
	\leq \lr \sigma^2 \; .
\end{align}
When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, assuming all $T(x)$ belong to the marginal polytope, we get
\begin{multline}
	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ||\hat \mu) }
	=\conj(\mu_*) \overbrace{- \E[\conj(\hat \mu)]}^{\leq - \conj(\E[\hat \mu])} \\
	+ \inv{n}\sum_i \E[\underbrace{\E[\langle  T(X_i) - \mu_*; \nabla\conj (\hat \mu) \rangle | X_j, j\neq i]}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance} } ]  \leq \frac{\sigma^2}{n}
\end{multline}
where we used the decomposition $\hat \mu = \inv{n}T(X_i) + (1-\inv{n}) \inv{n-1} \sum_{j\neq i}T(X_j)$ to apply~\eqref{eq:hypvariance}. 
In words, this variance assumption on $T(X)$ and $\conj$ immediately gives us a bound on the suboptimality. We can also apply this result to the MAP estimate $\hat \mu = \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n}$ but I did not manage to reach a satisfying conclusion about the bias. 
Note $\tilde \mu = \E[\hat \mu] = \frac{n_0 \mu_0 + n \mu_*}{n_0+n}$ and $\gamma_0= \frac{n_0}{n_0+n}$ and the step-size $\gamma = \inv{n_0 + n}$ .
\begin{align}
	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ||\hat \mu) } 
	&\leq \conj(\mu_*) - \conj(\tilde \mu) + \gamma_0 \langle \mu_0 - \mu_* ; \E[\hat \natp] \rangle + (1 - \gamma_0) \gamma \sigma^2  \\
	& = \bregmanconj (\mu_* ; \tilde \mu) - \langle \tilde \mu - \mu_* ;    \tilde \natp - \E[\hat \natp] \rangle + \frac{n}{(n_0+n)^2}\sigma^2
\end{align}
so we recover the $O(n^{-1})$ for the variance and we are still looking for $O(n^{-2})$ rate for the bias. The Bregman term in the bias should be in this spirit, given a quadratic approximation
\begin{align}
	 \bregmanconj (\mu_* ; \tilde \mu)  
	 = \bregmanconj (\mu_* ; \mu_* + \gamma_0 (\mu_0 - \mu_*) ) \approx \half \gamma_0^2 \norm{\mu_0 - \mu_*}^2_{\mSigma_*^{-1}}
\end{align}
where $\mSigma_* = \Cov_{\mu_*}(T(X))$. \RLP{beware of the order, this might be inaccurate}
However the scalar product is harder to bound. Hopefully it could be positive. If not we know that $\tilde \mu - \mu_* = \frac{n_0}{n_0+n} (\mu_0 - \mu_*) = O(n^{-1})$ , and we can hope that the same holds for $\tilde \natp - \E[\hat \natp] = \nabla \conj(\E[\hat \mu]) - \E[\nabla\conj (\hat \mu)]$, which kinda measures the non-linearity, or the curvature of $\conj$. Using a simple cauchy-schwartz, and the hessian of the conjugate, we might be able to get something.

This article \url{http://davidpfau.com/assets/generalized_bvd_proof.pdf} gives a bias variance decomposition which is applicable here.
